Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-23 articles/cycle/response-referee3.tex

    \referee{\textbf{p.~7, l.~14:}
        You keep mentioning the number of processors. I find this information
        slightly irrelevant, and it will soon become outdated as processing
        speed and models increase (GPUs for example). The only way it will
        serve the community is if there is a full description of the computer
        set up. For example, it could be that the simulations took 4 days on
        144 processors, but the processors were slow. Suggest removing these
        references.}

    We removed mentions of computing times and numbers of processors.

    \referee{\textbf{p.~10, l.~14:}
        This is a more general point. You eventually choose the EPICA record,
        and for justifiable reasons based on comparison to reconstructed ice
        extent and timing.  However, this is likely coincidence. EPICA is
        likely a complex record containing global and local antarctic
        influences upon climate. The ``real'' climate over the alps during
        glaciation is like decoupled from that of Antarctica to an extent.
        Therefore, the match you find is not an inference about climate, as
        different combinations of offsets may have made the same result
        (smoothed GRIP to remove some of the D-O scale noise?). You should make
        this explicit somewhere in the manuscript.}

    We agree. The following text was added in the conclusions:

    \msquote{%
        This suprising result is likely coincidental as there is no direct link
        between European and Antarctic climate. This highlights the need for
        more quantitative reconstructions of European palaeoclimate.}

@@

    This finding was better highlighted in the abstract (see our previous
    comment). However, we feel it is already well expressed by the following
    bullet point in the conclusion:

    \msquote{%
        The LGM (maximum) extent was a transient stage in which glaciers were
        out of balance with the contemporary climate. Its timing potentially
        varied across the range due to inherent glacier dynamics (Sect. 4.3).}

    \referee{\textbf{p.~17, l.~22:}
        Is the model recreating possible surging? Or would this be an
        over-interpretation given the uncertainty in climate and physics. Seems
        to fit with the enthalpy model of Benn and others for surging. As
        reads, it suggests that these areas were possible palaeo-surges, please
        clarify.}

    Our simulations reproduce a thermodynamical instability akin to a surge.
    The internal energy (so-called entropy) conservation scheme used in PISM
    allows for variations in ice temperature and liquid water content
    \citep{Aschwanden.etal.2012}, which may result in thermodynamical
    oscillations \citep{Pelt.Oerlemans.2012, Feldmann.Levermann.2017}.

    Unlike periodic surges \citep[as described by, e.g.,][]{Sevestre.Benn.2015},
    we interpret the model results as triggered by an initial perturbation of
    mass (and energy) caused by rapid climate cooling and inefficient mass
    transfer to lower elevations. We have rephrase the following sentence:

    \msquote{%
        Several Alpine lobes surge and overshoot their balanced extent before
        thinning and receding towards the mountains as they warm towards
        thermodynamical equilibrium.}

@@

    Actually we mean geologically recorded asymmetry. This clearly needed
    clarification. The corresponding sentence now reads:

    \msquote{%
        The observed asymmetric extent of ice north and south of the Alps can
        be explained by the modelled transient nature of the LGM extent without
        involving north-south gradients in temperature and precipitation change
        (Sect.~4.1).}

    \referee{%
        Additional references: Ballantyne, C.K. and Stone, J.O., 2015.
        Trimlines, blockfields and the vertical extent of the last ice sheet in
        southern Ireland. Boreas, 44(2), pp.277- 287. Kuchar, J., Milne, G.,
        Hubbard, A., Patton, H., Bradley, S., Shennan, I. and Edwards, R.,
        2012. Evaluation of a numerical model of the British–Irish ice sheet
        using relative sea-level data: implications for the interpretation of
        trimline observations.  Journal of Quaternary Science, 27(6),
        pp.597-605.}

% ----------------------------------------------------------------------
% References

\bibliographystyle{copernicus}
\bibliography{../../references/references}

### 2018-07-23 articles/cycle/alps-cycle.tex
Address referee 3 comment on surge.

    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Adige Glaciers, several thousand model years are needed to attain
    a thermodynamical equilibrium between cold-ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of the subglacial bedrock, that has been previously
    cooled downed by subfreezing air temperature before glacier advance.
    For larger glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. This is why several Alpine lobes overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium (supplementary animation). In
    contrast, peripheral ice caps with basal topography restricted to high
    elevations experience very low shear strain and virtually no basal sliding.
    They advance regularly on a frozen bed during the entire cold period,
    resulting in a late maximum stage (Fig.~\ref{fig:timing}).

>>>>

    For large Alpine glaciers flowing in deep valleys, such as the Rhone,
    Rhine and Adige Glaciers, several thousand model years are needed to attain
    a thermodynamical equilibrium between cold-ice advection from the high
    accumulation areas, upward diffusion of geothermal heat, and heat release
    from strong basal shear strain. Temperature evolution is, in part, limited
    by the slow warming of the subglacial bedrock, that has been previously
    cooled downed by subfreezing air temperature before glacier advance.
    For larger glaciers, this thermodynamical equilibrium is typically not yet
    attained around 25\,ka. Several Alpine lobes surge and overshoot their
    balanced extent before thinning and receding towards the mountains as they
    warm towards thermodynamical equilibrium (supplementary animation). In
    contrast, peripheral ice caps with basal topography restricted to high
    elevations experience very low shear strain and virtually no basal sliding.
    They advance regularly on a frozen bed during the entire cold period,
    resulting in a late maximum stage (Fig.~\ref{fig:timing}).
